Conservative Model Order Reduction for Fluid Flow

### 2018-10-05 sections/1.intro.tex
corrections added

Large scale simulations of fluid flows arise in a wide range of disciplines and industries. Therefore, MOR of fluid flows, specially when advective terms are dominant, is important. It is well known that conservation of the energy, specially kinetic energy, is essential for a qualitatively correct numerical integration of fluid flows. Conventional model reduction techniques often violates conservation of mass, momentum \cite{carlberg2018conservative}, or energy in fluid flows which result in an unstable reduced system, in particular for long time-integration. The method discussed in \cite{carlberg2018conservative} conserves the mass and momentum for a finite volume discretization scheme. However, a method that also conserves the energy of the fluid flow is not known to the authors.

Skew-symmetric formulation of fluid flows constructs a skew-symmetric differential operator, acting on the momentum vector field, that ensures conservation of quadratic invariants, such as energy. Combined wirh centered time and space discretization schemes, typically a finite differences discretization method, they recover time-symmetries of a fluid at the discrete level. Such discretization schemes are studies comprehensively over the past few decades and can be found in the works of \cite{morinishi2010skew,morinishi1998fully,desjardins2008high,tadmor1984skew,reiss2014conservative} and the references therein.

In this paper we discuss how to preserve skew-symmetry of the differential operators at the level of the reduced system. This results in conservation of quadratic invariants. Furthermore, we show that the reduced system, as a system of coupled differential equations, contains quadratic invariants and an associated energy which approximates the energy of the high-fidelity system. Therefore, a proper time stepping scheme preserves the reduced representation of the energy, and therefore, the loss in energy due to model reduction remains constant in time. Furthermore, we demonstrate, through numerical experiments, that a quasi-skew-symmetric form of fluid flow, i.e. a formulation where only spacial differential operators are in a skew-symmetric form, offer remarkable stability properties in terms of MOR. This allows an explicit time-integration to be utilized while recovering robustness of skew-symmetric forms at the reduced level.

>>>>

Large scale simulations of fluid flows arise in a wide range of disciplines and industries. Therefore, MOR of fluid flows, specially when advective terms are dominant, is important. It is well known that conservation of the energy, specially kinetic energy, is essential for a qualitatively correct numerical integration of fluid flows. Conventional model reduction techniques often violates conservation of mass, momentum \cite{carlberg2018conservative}, or energy in fluid flows which result in an unstable reduced system, in particular for long time-integration. \edit{In \cite{barone2009stable} an entropy stable model reduction method for linear compressible flows is presented by considering an entropy-stable formulation of linearized compressible flows. Furthermore, a conservative model reduction for finite-volume models is presented in \cite{carlberg2018conservative} that conserves any quantity conserved by the finite-volume scheme. This method finds a reduced linear subspace that ensures conservation of quantities by solving an optimization problem with, generally nonlinear, equality constraints. The constrained optimization problem is solved online, and is only slightly more expensive than solving the unconstrained optimization problem associated with a typical Galerkin MOR.}

%\edit{The method discussed in \cite{carlberg2018conservative} conserves the mass and momentum for a finite volume discretization scheme. However, a method that also conserves the energy of the fluid flow is not known to the authors.}

Skew-symmetric formulation of fluid flows constructs a skew-symmetric differential operator, acting on the momentum vector field, that ensures conservation of quadratic invariants, such as energy. \edit{Combined with} centered time and space discretization schemes, typically a finite differences discretization method, they recover time-symmetries of a fluid at the discrete level. Such discretization schemes are studies comprehensively over the past few decades and can be found in the works of \cite{morinishi2010skew,morinishi1998fully,desjardins2008high,tadmor1984skew,reiss2014conservative} and the references therein.

In this paper we discuss how to preserve skew-symmetry of the differential operators at the level of the reduced system. This results in conservation of quadratic invariants. \edit{The conservation of quantities in the proposed method is guaranteed through the mathematical formulation of the reduced system, for any orthonormal reduced basis. Therefore, the offline and online computational costs of this method is comparable with conventional MOR techniques. However, other conservative model reduction methods, e.g. \cite{carlberg2018conservative}, often require solving multiple nonlinear optimization problems to ensure conservation which can increase the computational costs.} Furthermore, we show that the reduced system, as a system of coupled differential equations, contains quadratic invariants and an associated energy which approximates the energy of the high-fidelity system. Therefore, a proper time stepping scheme preserves the reduced representation of the energy, and therefore, the loss in energy due to model reduction remains constant in time. Furthermore, we demonstrate, through numerical experiments, that a quasi-skew-symmetric form of fluid flow, i.e. a formulation where only spacial differential operators are in a skew-symmetric form, offer remarkable stability properties in terms of MOR. This allows an explicit time-integration to be utilized while recovering robustness of skew-symmetric forms at the reduced level.

### 2018-10-05 sections/3.skew.tex
corrections added

\subsection{Conservation Laws} \label{sec:skew.1}
In the context of fluid flows, transport of conserved quantities, can be expressed as
\begin{equation} \label{eq:3.1}
	\frac{\partial }{\partial t} \rho \varphi + \nabla \cdot ( \rho u \varphi  ) = \nabla \cdot F_{\varphi}\quad \text{defined in} \quad \Omega \subset \mathbb R^{d}.
\end{equation}
Here, $d = 1,2$ or $3$, $\rho:\Omega\to \mathbb R$ is the density, $u\in \Omega \to \mathbb R^{d}$ is the velocity vector field, $\varphi$ is a measured scalar quantity of the flow, and $F_{\varphi}$ is the flux function associated to $\varphi$. Integration of \eqref{eq:3.1} over $ \Omega$ yields
\begin{equation} \label{eq:3.2}
	\frac{d}{dt} \int_{\Omega} \rho \varphi \ dx = \int_{\partial \Omega} (F_{\varphi} - \rho u \varphi) \cdot \hat n\ ds,
\end{equation}
where $\partial \Omega$ is the boundary of $\Omega$, and $\hat n$ is the unit outward normal vector to $\partial \Omega$. When the right hand side of \eqref{eq:3.2} vanishes, e.g. in case where $\Omega$ is periodic and $\varphi$ is flux-free, the quantity $(\rho \varphi)$ is conserved. Therefore, \eqref{eq:3.1} is referred to as the \emph{conservative form} and the convective term in \eqref{eq:3.1} is referred to as the \emph{divergence form}. However, using the \emph{continuity equation}
\begin{equation} \label{eq:3.3}
	\frac{\partial }{\partial t} \rho + \nabla \cdot (\rho u) = 0,
\end{equation}
we can rewrite \eqref{eq:3.1} as
\begin{equation} \label{eq:3.4}
	\rho \frac{\partial }{\partial t} \varphi + (\rho u)\cdot \nabla \varphi = \nabla \cdot F_{\varphi}.
\end{equation}
The convective term in this formulation is referred to as the \emph{advective form}. The \emph{skew-symmetric} form of the convective term is obtained by the arithmetic average of the divergent and the advective form:
\begin{equation} \label{eq:3.5}
	\frac{1}{2} \left( \rho \frac{\partial }{\partial t} \varphi + \frac{\partial }{\partial t} (\rho \varphi) \right) + \frac 1 2 \left( (\rho u)\cdot \nabla \varphi + \nabla \cdot (\rho u \varphi) \right) = \nabla \cdot F_{\varphi}.
\end{equation}
Multiplying \eqref{eq:3.5} with $\varphi$ yields
\begin{equation}
	\frac{\partial }{\partial t} \rho \varphi^2 + \nabla \cdot ( \rho u \varphi^2  ) = \varphi \nabla \cdot F_{\varphi}.
\end{equation}
Therefore, $\varphi^2$ is a conserved quantity for a flux-free $\varphi$. Since the divergence, the advective and the skew-symmetric forms are identical at the continuous level, $\varphi^2$ is a conserved quantity for all forms. However, the equivalence of these forms is not preserved through a general discretization scheme and we can not expect $\varphi^2$ to be a conserved quantity at the discrete level. To motivate numerical advantages of the skew-symmetric form consider the operator 
\begin{equation}
	S_{\rho u}(\cdot) = \frac 1 2 ( [ \nabla \cdot \rho u ] + (\rho u)\cdot \nabla )(\cdot).
\end{equation}
With a proper set of boundary condition, this operator is a skew-adjoint operator on $L^2$. Here, $[\cdot]$ indicates that the inside of the brackets act as a differential operator. This skew-adjoint property is used later to show the conservation of some quadratic quantities in \eqref{eq:3.1}. Similarly, we can define a skew-adjoint operator with respect to the time variable as
\begin{equation}
	S_{\rho,\partial_t} = \frac{1}{2} \left( \rho \frac{\partial}{\partial t} + [ \frac{\partial}{\partial t} \rho] \right).
\end{equation}
Here, the subscript $\partial t$ is to emphasize that $S_{\rho,\partial_t}$ is a differential operator with respect to $t$. A proper time and space discretization of $S_{\rho u}$ and $S_{\rho,\partial_t}$ can preserve the skewness property.

Numerical time integration of \eqref{eq:3.5} can be challenging since the time differentiation of different variables is present. Following \cite{morinishi2010skew}, we rewrite \eqref{eq:3.5} as
\begin{equation} \label{eq:3.6}
	\sqrt{\rho } \frac{\partial }{\partial t} (\sqrt \rho \varphi ) + S_{\rho u}(\varphi) = \nabla \cdot F_{\varphi}.
\end{equation}
Time integration of this form is presented in \cite{morinishi2010skew,reiss2014conservative}. Note that one can also generate a quasi-skew-symmetric form \cite{blaisdell1991numerical,morinishi2003dns} of \eqref{eq:3.1} as
\begin{equation} \label{eq:3.7}
	\frac{\partial }{\partial t} (\rho \varphi) + S_{\rho u}(\varphi) = \nabla \cdot F_{\varphi}.
\end{equation}
Even though this is not a fully skew-symmetric form (skew-symmetric only in space), the numerical stability of this form is significantly better than the divergence and advective form \cite{morinishi2010skew,blaisdell1991numerical,morinishi2003dns}. Note that this quasi-skew-symmetric form is identical to the skew-symmetric form in the incompressible limit.

@@

\subsection{Time integration}
Following \cite{reiss2014conservative,morinishi2010skew} we can construct a fully discrete second order accurate scheme for \eqref{sec:skew.3} as
\begin{equation} \label{eq:3.27}
	\left\{
	\begin{aligned}
	&\frac 1 2 \sqrt{\mathbf r} ^{n+1/2} \frac{\sqrt{\mathbf r}^{n+1} - \sqrt{\mathbf r}^{n}}{\Delta t} + DU^{n+1/2} \mathbf r^{n} = 0, \\
	& \sqrt{\mathbf r} ^{n+1/2}  \frac{\sqrt{ \mathbf R}^{n+1} u^{n+1} - \sqrt{\mathbf R}^{n}u^n}{\Delta t} + S_{\mathbf r^{n} \mathbf u^n} \mathbf u^{n+1/2}_\alpha + D \mathbf p^{n} = DT^{n}, \\
	& \frac 1 {\gamma -1} \frac{\mathbf p^{n+1} - \mathbf p^n}{\Delta t} + \frac{\gamma}{\gamma -1} D U^{n} \mathbf p - U^{n} D \mathbf p = - U^{n}D T^{n} + D (U^nT^n - \phi^n).
	\end{aligned}
	\right.
\end{equation}
Here, $\Delta t$ is the time step, superscript $n$ denotes evaluating at $t = n\Delta t$, superscript $n+1/2$ denotes the arithmetic average of a variable evaluated at $t=n\Delta t$ and $t=(n+1)\Delta t$, the square root sign denotes element-wise application of square root, and 
\begin{equation}
	\mathbf u_{\alpha}^{n+1/2} = \frac{\sqrt{\mathbf R}^{n+1} \mathbf u^{n+1} + \sqrt{\mathbf R}^{n} \mathbf u^{n}}{2\sqrt{\mathbf r}^{n+1/2} }.
\end{equation}
As discussed in \cite{reiss2014conservative}, this time discretization scheme preserves the symmetries expressed in \eqref{eq:3.17}, \eqref{eq:3.24}, \eqref{eq:3.25}, and \eqref{eq:3.26}. In the incompressible case, the method reduces to the implicit mid-point scheme \cite{hairer2006geometric}. For further information see \cite{reiss2014conservative,morinishi2010skew}.

>>>>

\subsection{Conservation Laws} \label{sec:skew.1}
In the context of fluid flows, transport of conserved quantities, can be expressed as
\begin{equation} \label{eq:3.1}
	\frac{\partial }{\partial t} \rho \varphi + \nabla \cdot ( \rho u \varphi  ) = \nabla \cdot F_{\varphi}\quad \text{defined in} \quad \Omega \subset \mathbb R^{d}.
\end{equation}
Here, $d = 1,2$ or $3$, $\rho:\Omega\to \mathbb R$ is the density, $u\in \Omega \to \mathbb R^{d}$ is the velocity vector field, $\varphi$ is a measured scalar quantity of the flow, and $F_{\varphi}$ is the flux function associated to $\varphi$. Integration of \eqref{eq:3.1} over $ \Omega$ yields
\begin{equation} \label{eq:3.2}
	\frac{d}{dt} \int_{\Omega} \rho \varphi \ dx = \int_{\partial \Omega} (F_{\varphi} - \rho u \varphi) \cdot \hat n\ ds,
\end{equation}
where $\partial \Omega$ is the boundary of $\Omega$, and $\hat n$ is the unit outward normal vector to $\partial \Omega$. \edit{This means that the quantity $(\rho\varphi)$ is explicitly conserved over control volumes.} Therefore, \eqref{eq:3.1} is referred to as the \emph{conservative form} and the convective term in \eqref{eq:3.1} is referred to as the \emph{divergence form}. However, using the \emph{continuity equation}
\begin{equation} \label{eq:3.3}
	\frac{\partial }{\partial t} \rho + \nabla \cdot (\rho u) = 0,
\end{equation}
we can rewrite \eqref{eq:3.1} as
\begin{equation} \label{eq:3.4}
	\rho \frac{\partial }{\partial t} \varphi + (\rho u)\cdot \nabla \varphi = \nabla \cdot F_{\varphi}.
\end{equation}
The convective term in this formulation is referred to as the \emph{advective form}. The \emph{skew-symmetric} form of the convective term is obtained by the arithmetic average of the divergent and the advective form:
\begin{equation} \label{eq:3.5}
	\frac{1}{2} \left( \rho \frac{\partial }{\partial t} \varphi + \frac{\partial }{\partial t} (\rho \varphi) \right) + \frac 1 2 \left( (\rho u)\cdot \nabla \varphi + \nabla \cdot (\rho u \varphi) \right) = \nabla \cdot F_{\varphi}.
\end{equation}
Multiplying \eqref{eq:3.5} with $\varphi$ yields
\begin{equation}
	\frac{\partial }{\partial t} \rho \varphi^2 + \nabla \cdot ( \rho u \varphi^2  ) = \varphi \nabla \cdot F_{\varphi}.
\end{equation}
\edit{Therefore, $\rho \varphi^2$} is a conserved quantity for a flux-free $\varphi$. Since the divergence, the advective and the skew-symmetric forms are identical at the continuous level, $\varphi^2$ is a conserved quantity for all forms. However, the equivalence of these forms is not preserved through a general discretization scheme and we can not expect $\varphi^2$ to be a conserved quantity at the discrete level. To motivate numerical advantages of the skew-symmetric form consider the operator 
\begin{equation}
	S_{\rho u}(\cdot) = \frac 1 2 ( [ \nabla \cdot \rho u ] + (\rho u)\cdot \nabla )(\cdot).
\end{equation}
With a proper set of boundary condition, this operator is a skew-adjoint operator on $L^2$. Here, $[\cdot]$ indicates that the inside of the brackets act as a differential operator. This skew-adjoint property is used later to show the conservation of some quadratic quantities in \eqref{eq:3.1}. Similarly, we can define a skew-adjoint operator with respect to the time variable as
\begin{equation}
	S_{\rho,\partial_t} = \frac{1}{2} \left( \rho \frac{\partial}{\partial t} + [ \frac{\partial}{\partial t} \rho] \right).
\end{equation}
Here, the subscript $\partial t$ is to emphasize that $S_{\rho,\partial_t}$ is a differential operator with respect to $t$. A proper time and space discretization of $S_{\rho u}$ and $S_{\rho,\partial_t}$ can preserve the skewness property.

Numerical time integration of \eqref{eq:3.5} can be challenging since the time differentiation of different variables is present. Following \cite{morinishi2010skew}, we rewrite \eqref{eq:3.5} as
\begin{equation} \label{eq:3.6}
	\sqrt{\rho } \frac{\partial }{\partial t} (\sqrt \rho \varphi ) + S_{\rho u}(\varphi) = \nabla \cdot F_{\varphi}.
\end{equation}
Time integration of this form is presented in \cite{morinishi2010skew,reiss2014conservative}. Note that one can also generate a quasi-skew-symmetric form \cite{blaisdell1991numerical,morinishi2003dns} of \eqref{eq:3.1} as
\begin{equation} \label{eq:3.7}
	\frac{\partial }{\partial t} (\rho \varphi) + \edit{ \frac 1 2 \left( \nabla \cdot(\rho u \varphi) + \rho u \cdot \nabla \varphi + \varphi \nabla \cdot(\rho u) \right) } = \nabla \cdot F_{\varphi}.
\end{equation}
Even though this is not a fully skew-symmetric form (skew-symmetric only in space), the numerical stability of this form is significantly better than the divergence and advective form \cite{morinishi2010skew,blaisdell1991numerical,morinishi2003dns}. Note that this quasi-skew-symmetric form is identical to the skew-symmetric form in the incompressible limit.

@@

\subsection{Time integration}
Following \cite{reiss2014conservative,morinishi2010skew} we can construct a fully discrete second order accurate scheme for \eqref{sec:skew.3} as
\begin{equation} \label{eq:3.27}
	\left\{
	\begin{aligned}
	&\frac 1 2 \sqrt{\mathbf r} ^{n+1/2} \frac{\sqrt{\mathbf r}^{n+1} - \sqrt{\mathbf r}^{n}}{\Delta t} + DU^{n+1/2} \mathbf r^{n} = 0, \\
	& \sqrt{\mathbf r} ^{n+1/2}  \frac{\sqrt{ \mathbf R}^{n+1} \mathbf u^{n+1} - \sqrt{\mathbf R}^{n}\mathbf u^n}{\Delta t} + S_{\mathbf r^{n} \mathbf u^n} \mathbf u^{n+1/2}_\alpha + D \mathbf p^{n} = DT^{n}, \\
	& \frac 1 {\gamma -1} \frac{\mathbf p^{n+1} - \mathbf p^n}{\Delta t} + \frac{\gamma}{\gamma -1} D U^{n} \mathbf p^n - U^{n} D \mathbf p^n = - U^{n}D T^{n} + D (U^nT^n - \phi^n),
	\end{aligned}
	\right.
\end{equation}
\edit{and all the variables are stored at the same location with the same resolution (colocated).} Here, \edit{$\mathbf R$ contains elements of $\mathbf r$ on its diagonal,} $\Delta t$ is the time step, superscript $n$ denotes evaluating at $t = n\Delta t$, superscript $n+1/2$ denotes the arithmetic average of a variable evaluated at $t=n\Delta t$ and $t=(n+1)\Delta t$, the square root sign denotes element-wise application of square root, and 
\begin{equation}
	\mathbf u_{\alpha}^{n+1/2} = \frac{\sqrt{\mathbf R}^{n+1} \mathbf u^{n+1} + \sqrt{\mathbf R}^{n} \mathbf u^{n}}{2\sqrt{\mathbf r}^{n+1/2} }.
\end{equation}
As discussed in \cite{reiss2014conservative}, this time discretization scheme preserves the symmetries expressed in \eqref{eq:3.17}, \eqref{eq:3.24}, \eqref{eq:3.25}, and \eqref{eq:3.26}. In the incompressible case, the method reduces to the implicit mid-point scheme \cite{hairer2006geometric}. For further information see \cite{reiss2014conservative,morinishi2010skew}.
